Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational register data from Sweden, England and Wales

### 2019-10-17 reports/article.Rmd
new draft sent to NH, AG, SN and OR.

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on Swedish and English register data"
author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Szilard Nemes^2,4^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: P:/library.bib
output: 
  bookdown::word_document2:
    toc: false
    df_print: kable
---

```{r setup, include = FALSE}
options(
  digits            = 0,
  knitr.kable.NA    = '',
  repos = list(CRAN = "https://cran.rstudio.com/")
)
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)

# Load all files
walk(file.path("../cache", dir("../cache", ".RData")), load, .GlobalEnv)
```

@@

**Correspondence:** 
AG: anne.l.garland@gmail.com, Phone: +46 498 26 80 00, Postal: Orthopedic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden  
EB: erik.bulow@registercentrum.se  
SN: szilard.nemes@registercentrum.se  
NH: nils.hailer@surgsci.uu.se

@@

**OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity. Diagnosis-based tools such as the Charlson or Elixhauser comorbidity indices have been used to quantify this. Those instruments are however complex and requires extensive data sets not routinely available in most doctor-patient settings. We searched for a simplified model to  substitute those complex indices in predicting early mortality after THA.

**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to the national population register, the National Patient Register and the longitudinal integration database for health insurance and labor market studies from Statistics Sweden. We used a bootstrap ranking procedure with logistic regression and a LASSO-type penalty to develop a prediction model for patient deaths within 90 days after surgery. Predictive power was assessed by the area under the receiver operating characteristic curve (AUC). The final model was applied to British data for external validation.

```{r}
get_auc <- function(mod) subset(brlasso_tbl_auc, Model == mod, "AUC", drop = TRUE)
```

**RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`. Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r get_auc("BRL any")`). This model was superior to the established but complex Charlson comorbidity index (AUC = `r get_auc("CCI")`), and the Elixhauser comorbidity score (AUC = `r get_auc("ECI")`). A web calculator to aid clinical usage was published at https://erikbulow.shinyapps.io/thamortpred/.

**CONCLUSION:** We found a relatively simple prediction model of 90 day mortality after total hip arthroplasty. This model requires less data and is easier to calculate compared to previously well-known comorbidity indices.

@@

The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA) [@Inacio2015], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016]. In research settings, comorbidity is commonly measured using multi-faceted diagnosis- or prescription-based coding algorithms [@Bozic2013]. @Inacio2016 studied the ability of the Charlson (CCI) and Elixhauser (ECI) Comorbidity indices to to predict mortality after THA and total knee arthroplasty. These comorbidity measures are quite complex to estimate and are based on the availability of extensive data sets including in- and outpatient data on ICD-codes. Oftentimes, such data sets can only be created by linking several population-based registries, raising both ethical and practical concerns. Also, each of the comorbidity indices exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009]. Interpretation and comparison between different studies is therefore difficult.

Comorbidity data have also been used in several universal and arthroplasty-specific risk prediction tools to make risk profiles for individual patients. In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016]. No model has so far been broadly accepted for elective THA however [@Manning2016; @Bulow2017]. An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice. We aimed to find such a model to predict the risk of 90-day postoperative mortality after elective THA.

@@

Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented primary hip osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were included in the development phase of the study (Figure \@ref(fig:flowchart)).
Only the last operated hip was accounted for in patients with bilateral THA [@Bulow2019a].

```{r flowchart, fig.cap = "Flowchart depicting inclusion criteria and number of patients included in the development phase of the model."}
knitr::include_graphics("../graphs/flowchart.png")
```

Data linkage, based on the unique identity numbers assigned to all Swedish residents [@Ludvigsson2009], were used to collect data from a variety of sources, as previously described by @Cnudde2016.

Age, sex, body mass index (BMI), ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from SHAR, with a completeness of 96-98 % [@shpr2018]. Data on education level (low/middle/high) and civil status (married/un-married/divorced/widow[er]), were collected from the longitudinal integration database for health insurance and labor market studies from Statistics Sweden [@Ludvigsson2019]. The Swedish National Patient Register was used for comorbidity data during one year before surgery. The register contains all relevant diagnoses coded by ICD-10, as well as admissions and discharge dates for in- and outpatient visits in all private and public hospitals [@Ludvigsson2011]. Death dates were linked from the national population register.

Comorbidity was recognized by individual ICD-10 grouped into 17 categories according to CCI [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to ECI [@Elixhauser1998; @Quan2005]. Patients with no recorded hospital visits during one year before surgery, were assumed to have no comorbidity.

```{r tabcategorization}
knitr::kable(
  tab_categorization, 
  caption = "Categorization of individual Charlson (CCI) and Elixhauser (ECI) comorbidities into broader comorbidities.")

```

Some comorbidities were identified by both CCI and ECI, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol). We used those categories to establish `r nrow(tab_categorization)` broader categories (Table \@ref(tab:tabcategorization)) in addition to 5 standalone ECI classes that were kept unchanged (`r glue::glue_collapse(ECI_keep_text, ", ", last = " and ")`). Groups were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data. Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process described below. The final model was also altered to not include cancer as a predictor. Patients with cancer are sometimes treated differently in the clinical setting, introducing difficulties in interpretation of this variable.

@@

We used the Kaplan-Meier estimator to assess unadjusted cumulative survival.

Further analysis were based on logistic regression since no censoring occurred within the 90 day study period. We used a modelling procedure described by @Guo2015 as a bootstrap ranking procedure with a logistic least absolute shrinkage and selection operator (LASSO) model. 
Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. 
1,000 bootstrap samples were drawn from the initial data set [@Austin2004]. We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. Absolute values from those estimates were used as a measure of variable importance. Piece-wise linear regression was then used to detect a break point where a significant drop in variable importance were observed. Potential predictors with variable importance above this break point were considered important and kept as model candidates. The whole process was repeated ten times.

Covariates that were selected each of the ten times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables. We will call this model "BRL all", where BRL stands for bootstrap ranking LASSO. A similar model, including any variable selected at least one out of ten times will be called "BRL any". Univariable models with the ASA score, CCI and ECI were used for comparison, as well as a multivariable model with age and sex. Each model including age where fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.

Each of those models were used to predict the probability of death within 90 days for each patient. Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predicitve power. Models with a lower 95 % confidence limit [@Delong1988] above 0.7, were considered good.

Odds ratios for the model with highest AUC were estimated with 95 % confidence intervals based on interpolations of profile traces [@Ripley2002].

We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr and pROC. We build an online web calculator using the shiny package. All R-scripts and necessary software (but no personal data) is available as a live Binder environment (https://mybinder.org/v2/gh/eribul/thamortpred/master?urlpath=rstudio). A static archived version is also available at zenodo.org/XXX.
**Prepared but non-public until paper is accepted/published!**

@@

Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (271-14 and 360-13).

@@

We found `r format(nrow(df), big.mark = ", ")` patients (Figure \@ref(fig:flowchart)), `r paste(range(df$P_Age), collapse = " - ")` years old, whereof `r round(mean(df$P_Sex == "Female") * 100)` % were female. `r with(df, sprintf("%d (%.2f %%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that. The unadjusted cumulative 90-day survival was `r surv90d`.

Some characteristics of the study population are presented in Table \@ref(tab:tab1). `r mean(df$CCI_index_quan_original != 0) * 100` % of all patients had at least one pre-surgery comorbidity according to CCI, and `r mean(df$ECI_index_sum_all  != 0) * 100` % according to ECI. The proportion of patients with ASA class 3 was `r mean(df$P_ASA == 3) * 100` %. Most individual comorbidities were more common among patients who died.

```{r tab1}
knitr::kable(
  table1, 
  caption = "Baseline demographics. CCI/ECI = Charlson/Elixhauser comorbidity indices. Comorbidities prefixed with ECI are defined by the Elixhauser classification. Remaining comorbidities are based on the previously described combination of CCI and ECI. Comorbidities recorded for at least one patient who survived 90 days, and one who did not, were modeled as potential predictors.")
```

There were `r nrow(comb_lgl)` comorbidities that were not recorded for any patient who died: `r comb_lgl_text`. Corresponding variables were excluded from the modelling process.

The "BRL any" model included `r coefs_print`. The "BRL all model only included `r coefs_print_all`.

There were no observed differences for two versus three knots in models with age modeled by restricted cubic splines. We will therefore only present results for three knots. No use of splines however improved the observed AUCs compared to simpler main effect models. The "BRL any" model had an estimated AUC (95 % CI) of `r get_auc("BRL any")`. The altered version without cancer performed equally good. The "BRL any" model was significantly better than the "BRL all" model with AUC `r get_auc("BRL all")`. Univariable models with ASA, CCI or ECI performed poorly with AUC lower than, or at least not distinguishable from 0.7. The model with age and sex performed better but still with AUC not significantly higher than 0.7 (Table \@ref(tab:auc)).

```{r auc, warning = FALSE}
knitr::kable(
  brlasso_tbl_auc, 
  caption = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to 'BRL all', a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices. Age was included as either a main effect, or in the form of restricted cubic splines (RCS) with three knots.")

```

ROC curves for some of the models are displayed in Figure \@ref(fig:rocs). The "BRL any" model is superior to all other models.

```{r rocs, fig.cap = "ROC curves for some of the models. The 'BRL any' model is distinguished from other models, which are partially over-lapping."}
knitr::include_graphics("../graphs/brlasso_roc.png")
```

AUCs and 95 % confidence intervals are illustrated for the same models in Figure \@ref(fig:aucci)).

```{r aucci, fig.cap = "Area Under the Curve (AUC) as a measure of predictive power for the 'BRL any' model compared to a simpler model with age and sex, as well as univariable models with ASA score and the Charlson (CCI) or Elixhauser (ECI) comorbidity indices."}

knitr::include_graphics("../graphs/brlasso_auc_ci.png")
```

The ability of model "BRL any" to estimate probabilities of death within 90 days is illustrated in Figure \@ref(fig:sep). Patients who died had, on average, higher predicted probabilities to do so.

```{r sep, fig.cap = "Patients who died within 90 days (blue) were, on average, estimated to have a higher probability to do so."}
knitr::include_graphics("../graphs/brlasso_separation_auc.png")
```

Estimated coefficients and corresponding odds ratios for the "BRL any" model is presented in Table \@ref(tab:brlany).

```{r brlany}

knitr::kable(
  select(brlasso_tbl_coefs, term, X = X_math, beta, OR, p),
  caption = 'Estimated coefficients and odds ratios with 95 % confidence intervals for the "BRL any" model. Notations from the X-column is used in the formula in the disussion section.',
  escape = FALSE
)
```

We have provided a web calculator to aid clinical model usage in practice (https://erikbulow.shinyapps.io/thamortpred/).

@@

In this nationwide cohort study we intended to compare the performance of a set of easily accessible data that are routinely collected in daily clinical practice with the complex comorbidity coding algorithms suggested by Charlson and Elixhauser. We found that a multivariable main effects logistic regression model with `r coefs_print` was able to make better predictions than either CCI or ECI.

The resulting "BRL any" model can predict the probability of death ($p$) within 90 days for new patients as: `r coefs_form` where $X_i$ indicate independent variables as notated in table \@ref(tab:brlany).
This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years).

```{r}
get_risk <- function(who) preds[preds$id == who, "p", drop = TRUE]
```

For example a `r min(df$P_Age)` yer old woman with ASA = 1 and none of the important comorbidities would only have a `r get_risk("baseline")` risk to die within 90 days after elective THA surgery. Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have a higher risk of `r get_risk("healthy_female")`. A man, `r quantile(df$P_Age, .75)` years old (the third age quantile) with ASA = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")` risk. The perhaps unrealistic case of a `r max(df$P_Age)` year old man with ASA = 3 and all listed comorbidities would have a theoretical risk as high as `r get_risk("high_risk")`. Note however that this extreme case relies on extrapolation which is highly unreliable, since no such person was actually observed.

Some covariates in the "BRL any" model were not statistically significant by themselves but were still relevant due to unobserved heterogeneity [@Mood2010]. Obesity for example is known to be associated with a higher risk of morbidity and all-cause mortality [@Must2000]. However, previous studies on primary THA cohorts have not indicated a higher risk of mortality in obese patients [@Wallace2014]. An explanation could be that obese patients selected for THA are comparably healthy and often younger.

Cancer could also be dropped as a predictor without any loss of predictive power. It was also encouraging that socio-demographic factors such as education and civil status, or organizational factors such as type of hospital, did not have a strong enough influence of death to be included in the final model.

It is known that male sex is associated with earlier deaths and that the remaining life span will decrease with increased age. It is less obvious that this relation must be linear. We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good. Our model includes ASA class which is routinely assessed pre-operatively in most developed countries. It is however known to have a high degree of internal variability [@Haynes1995]. It has previously been compared to the CCI, but not with respect to mortality after THA [@Whitmore2014; @Kork2015]. Patients with ASA 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals. It can be questioned whether such classification is correct for our cohort. Comorbidities are also known to influence the outcome after THA [@Inacio2015; @Gordon2013; @Hofstede2016]. Coding algorithms on the other hand are complex and not used in clinical settings since the administrative burden is too high. CCI comprise 1,178 ICD-10 codes and ECI 1,516. Therefore, such indices are only used by researchers.

Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anesthesia management. A number of risk prediction tools of various complexity for adverse outcomes after total joint replacements have been introduced but none has been broadly accepted [@Manning2016]. In the context of trauma surgery outcome prediction tools are common, and it seems possible to reduce the number of items without losing predictive power [@Gerdin2016]. Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.

A strength of this study is the nationwide design with a large cohort of primary THA patients. We were able to use exact data linkage by the Swedish unique identity numbers and had no censoring. Our data sources are valid with low proportions of missing data [@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011].

The risk of coding errors might be a limitation to the study, especially so if coding routines might change over time. It should also be remembered that the risk model does not study THA as an observed intervention. We merely followed the cohort who did already have THA. Hence, deaths within 90 days might occur for the patients regardless if THA is inserted or not. The proximity in dates however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the deaths observed. The insertion of an elective THA is always preceded by a clinical judgement. Hence, no patient with a foreseen death near-by is given THA to begin with. We therefore believe that at least a non-significant proportion of deaths within 90 days are related to the THA surgery, or with complications thereafter.

@@

AG and NH initiated the study and managed the ethical review board application. EB, EL and SN performed the statistical analyses. AG and EB drafted the manuscript. All authors edited and finalized the manuscript.

>>>>

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational register data from Sweden, England and Wales"
author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: P:/library.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

```{r setup, include = FALSE}
options(
  digits            = 2,
  knitr.kable.NA    = '',
  repos = list(CRAN = "https://cran.rstudio.com/")
)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)

# Load all files
walk(file.path("../cache", dir("../cache", ".RData")), load, .GlobalEnv)
source("../lib/clean_names.R")
```

@@

**Correspondence: [OR EB?]** 
anne.garland@surgsci.uu.se, Phone: +46 498 26 80 00, Postal: Orthopedic Department, Visby Hospital, St Goransgatan 10, 621 84 Visby, Sweden

##### PAGE BREAK

@@

**OBJECTIVE:** Shared decision making is essential before total hip arthroplasty (THA) surgery. Comorbidities are associated with an increased risk of 90-day morbidity and mortality, but diagnosis code-based instruments such as the Charlson or Elixhauser comorbidity indices are impossible to use in a clinical setting, and the American Society for Anesthesiologists (ASA) grade is imprecise. We searched for a simple model to predict early mortality after THA.

**PATIENTS AND METHODS:** We studied `r format(nrow(df), big.mark = ",")` Swedish patients operated with a cemented THA due to primary osteoarthritis and linked data from several national registries in order to collect information on demographics and comorbidities. Bootstrap ranking procedures were used together with logistic LASSO regression to develop prediction models for death within 90 days. Predictive accuracy was assessed by the areas under the receiver operating characteristic curve (AUC). The final model was externally validated on a data-set from the National Joint Registry of England and Wales .

```{r}
get_auc <- function(mod) {
  subset(brlasso_tbl_auc, Model == mod, "AUC", drop = TRUE) %>% 
    {gsub("(", "(95% CI: ",., fixed = TRUE)}
}
```

**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d`. Best predictive performance for 90-day mortality was found for a model combining `r coefs_print_reduced` (AUC = `r get_auc("BRL reduced (age as main effect)")`). This model was superior to the  Charlson (AUC = `r get_auc("Charlson")`) and Elixhauser (AUC = `r get_auc("Elixhauser")`) comorbidity measures. A web calculator to aid clinical usage was published at https://erikbulow.shinyapps.io/thamortpred/ **[Would it be possible to embedd it at shpr.se instead?]**.

**CONCLUSION:** We found a relatively simple prediction model of 90-day mortality after THA. This model requires less data and is easier to use compared to previously well-known comorbidity indices.

##### PAGE BREAK

@@

Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process. In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure, but the risk of adverse events in the short and long term is an essential part of preoperative discussions and various risk prediction models have been developed.[@Price2019]

90-day mortality after THA surgery performed due to osteoarthritis is low, ranging between 0.2% and 0.6% in large joint registry analyses where adjustment for comorbidities is possible.[@Hunt2013; @Garland2017]

The presence of comorbidities is associated with a shorter remaining life span, but the Charlson and Elixhauser comorbidity indices poorly predict mortality after THA and total knee arthroplasty.[@Bulow2017; @Inacio2015] Additionally, these complex comorbidity instruments are based on the availability of extensive data-sets including in- and outpatient data coded by the 10th revision of the international classification of diseases (ICD-10), and each of the comorbidity indices exists in numerous versions.[@Brusselaers2017; @Walraven2009] Interpretation and comparison of different studies is therefore difficult.

Comorbidity data are used to construct universal or arthroplasty-specific risk prediction tools, but no model has so far been broadly accepted to predict mortality after  elective THA.[@Manning2016; @Bulow2017] In order to propel arthroplasty surgery into the age of precision medicine, easily applicable tools to predict short- and long-term complications are urgently needed. We aimed to define a model that accurately predicts the risk of 90-day postoperative mortality after THA.

@@

Patients recorded in the Swedish Hip Arthroplasty Register (SHAR) with cemented THA due to osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` were used to develop, and internally validate, the prediction model (Figure \@ref(fig:flowchart)). Data from the National Joint Register in England and Wales (NJR) was later used for external validation. 
Only the last operated hip was accounted for in patients with bilateral THA. [@Bulow2019a]

Data linkage, based on the 10-digit identity numbers assigned to all Swedish residents,[@Ludvigsson2009] were used to collect data from a variety of sources, as previously described.[@Cnudde2016]

Age, sex, body mass index (BMI), the American Society for Anesthesiologists (ASA) grade, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, which has a completeness of 96-98%.[@shpr2018] Data on education level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) and civil status (married/un-married/divorced/widow[er]), were collected from the longitudinal integration database for health insurance and labor market studies from Statistics Sweden.[@Ludvigsson2019] The Swedish National Patient Register was used to assess comorbidities during the year preceding index surgery. This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals.[@Ludvigsson2011] Death dates were linked from the national population register.

Comorbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005]

Some comorbidities were identified by both Charlson and Elixhauser, and some distinct comorbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol). We therefore combined individual categories to establish `r nrow(tab_categorization)` broader categories of comorbidity (Table \@ref(tab:tabcategorization)). Conditions were merged according to clinical relevance as to be recognized in a patient-doctor meeting without access to external register data.

@@

We used the Kaplan-Meier estimator to assess unadjusted mortality.

Further analysis were based on logistic regression since no censoring occurred within the 90 day study period. We used a modelling procedure with bootstrap ranking and logistic least absolute shrinkage and selection operator (LASSO) model.[@Guo2015] 
Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1.

Comorbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. Absolute values from those estimates were used as a measure of variable importance. Piece-wise linear regression was  used to detect a break point where a significant drop in variable importance were observed. Potential predictors with variable importance above this break point were considered important and kept as model candidates. The whole process was repeated 100 times.

Covariates that were selected at least one out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables. We will call this model "BRL" for bootstrap ranking LASSO. A reduced model with variables chosen at least 33 out of the 100 times, is called "BRL reduced". This model, but without cancer as a predictor, was also evaluated, since patients with cancer are sometimes treated differently [NEEDS CLARIFICATION!].

Univariable models with the ASA grade, Charlson or Elixhauser comorbidity indices were used for benchmarking, as well as a multivariable model with age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots.

Each of those models were used to predict the probability of death within 90 days for each patient. Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of predictive accuracy. Models with a lower 95% confidence limit above 0.7, were considered good. Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples for internal validation. The bias-corrected Somers' $D_{xy}$ rank correlation was used to adjust those intervals for optimism based on 100 re-samples.[@Miller1991] Calibration bands were made to graphically assess model calibration.[@Nattino2016]

Odds ratios for the final model were estimated with 95% confidence intervals based on interpolations of profile traces.[@Ripley2002]

We also built an online web calculator available at https://erikbulow.shinyapps.io/thamortpred/. Patients can use it to easily estimate their own risk of mortality. This estimate can be used in a discussion of shared decision making on whether to operate or not.

We used `r substr(R.version.string, 1, 15)` (R Foundation for Statistical   Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC and shiny. All R-scripts and necessary configurations (but no personal data) is available at zenodo.org/XXX. A linked Binder environment is also available for interactive usage.

@@

Ethical approval for this study was obtained from the Regional Ethical Review Board in Gothenburg (271-14 and 360-13).
**[COI?]**

@@

After application of exclusion criteria, `r format(nrow(df), big.mark = ",")` patients (age `r paste(range(df$P_Age), collapse = " - ")`, `r round(mean(df$P_Sex == "Female") * 100)`% females) were included in the analysis (Figure \@ref(fig:flowchart)). `r with(df, sprintf("%d (%.2f%%)", sum(death90), mean(death90) * 100))` patients died within 90 days and no one was censored before that. The unadjusted risk of 90 day mortality was `r surv90d`.

Characteristics of the study population are presented in Table \@ref(tab:tab1). `r round(mean(df$CCI_index_quan_original != 0) * 100)`% of all patients had at least one pre-surgery comorbidity according to Charlson and `r round(mean(df$ECI_index_sum_all  != 0) * 100)`% according to Elixhauser. The proportion of patients with ASA grade 3 was `r round(mean(df$P_ASA == 3) * 100)`%. Most comorbidities were more common among patients who died.

There were `r nrow(comb_lgl)` comorbidities that were not recorded for any patient who died: `r comb_lgl_text`, and these variables were thus excluded from the modelling process.

The "BRL" model included `r coefs_print`. The "BRL reduced" model only used `r coefs_print_reduced` (Table \@ref(tab:brlprop)).

We observed no relevant differences between models based on age described by restricted cubic splines with two versus three knots and therefore only present results for models based on three knots. The use of splines did however not improve the accuracy of the models when compared to simpler main effect models. The "BRL" model had an estimated AUC of `r get_auc("BRL (age as main effect)")`, and the reduced version performed equally good with AUC `r get_auc("BRL reduced (age as main effect)")`. Univariable models with ASA grade, Charlson or Elixhauser performed poorly with an AUC that was lower than 0.7. A model based on only age and sex performed slightly better but still with an AUC not statistically significantly above 0.7 (Table \@ref(tab:auc) and Figure \@ref(fig:rocs)).

AUCs and 95% confidence intervals, with and without adjustments for optimism, are illustrated for some models in Figure \@ref(fig:aucci)).

The ability of the "BRL reduced" model to estimate probabilities of death within 90 days is illustrated in Figure \@ref(fig:sep). Patients who died had, on average, higher predicted probabilities to do so. The model calibration was good for probabilities up to 3%, while higher probabilities might be over-estimated (Figure \@ref(fig:calibration)).

Estimated coefficients and corresponding odds ratios for the "BRL reduced" model is presented in Table \@ref(tab:brlreduced).

@@

We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` was able to better predict 90-day mortality than more complex models such as the Charlson or Elixhauser comorbidity indices.

The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(\beta X)]$ where `r coefs_form`
This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years) with observed combinations of diseases.

```{r}
get_risk <- function(who) people[people$id == who, "p", drop = TRUE]
```

For example a `r min(df$P_Age)` year old woman with ASA grade = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery. Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have an elevated risk of `r get_risk("healthy_female")`. A `r quantile(df$P_Age, .75)` years old man (the third age quantile) with ASA grade = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")`. If that man was instead 99 years old (the maximum observed age), with cancer as well, his risk would be `r get_risk("sickest_man")`. Note however that observed covariate patters with predicted probabilities above 3% were rare. Only `r mean(pred_probs > 3) * 100`% (`r sum(pred_probs > 3)` patients) had those (Figure \@ref(fig:sep)). Estimated mortality risks above 3% are therefore subject to extrapolation. Some risk calculators ignore this problem,[@Bozic2013a] but we think this should be acknowledged.

Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obisity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the national patient register (`r mean(df$c_obesity) * 100`%). This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`. Thus, obesity might be indicated in the risk calculator only if the condition is severe.

The "BRL reduced" model without cancer had the same AUC as a model with cancer. The AUC value itself is a rank-based and scale-invariant measure that does not account for model calibration. Both models were however equally well calibrated as well, for estimated probabilities within the recommended range (0-3%). The model without cancer however performed worse for probabilities above 3%.

It is known that male sex is associated with earlier deaths and that the remaining life span will decrease with increased age. It is less obvious that this relation must be linear. We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good. Our model includes ASA grade, which is routinely assessed pre-operatively in most developed countries. It is however known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson comorbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA grade 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals. Patients with those conditions are unlikely to get elective THA.

Comorbidity burden is also known to influence the outcome after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017] Coding algorithms on the other hand are complex and not used in clinical settings since the administrative burden is too high. The Charlson classification comprise 1,178 ICD-10 codes and the Elixhauser 1,516. Those indices are therefore only used by researchers.

Risk prediction may be useful in the process of patient selection prior to surgery, in the preoperative risk management including a review of current medications, and in perioperative anesthesia management. A number of risk prediction tools of various complexity for adverse events after total joint replacements have been introduced but none has been broadly accepted.[@Manning2016; @Price2019] In the context of trauma surgery, outcome prediction tools are common, and it seems possible to reduce the number of items without losing predictive accuracy.[@Gerdin2016] Our results indicate that the risk of early postoperative mortality after THA could be assessed by a relatively simple prediction model.

A strength of this study is the nationwide design with a large cohort of primary THA patients. We were able to use exact data linkage by the Swedish identity numbers and had no censoring. Our data sources are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011]

The risk of coding errors might be a limitation to the study, especially so if coding routines change over time. It should also be remembered that the risk model does not study THA as an observed intervention. We merely followed the cohort who did already have THA. Hence, deaths within 90 days might occur for the patients regardless if THA is inserted or not. The proximity in time however, the maximum of 90 days from THA to death, is an indication that the operation might have influenced the deaths observed. The insertion of an elective THA is always preceded by a clinical judgement. Hence, no patient with a foreseen death near-by is given THA to begin with. We therefore believe that at least a non-significant proportion of deaths within 90 days are related to  THA.

@@

**[ADD NJR CONTRIBUTIONS!]** AG and NH initiated the study and managed the ethical review board application. EB developed the statistical model. EL and AS performed external validation with data from NJR. AG and EB drafted the manuscript. All authors edited and finalized the manuscript.

# Acknowledgement

We would like to thank Szilard Nemes, previous senior statistician at the Swedish Hip Arthroplasty Register, for involvement in planning and interpreting the study.

##### PAGE BREAK

```{r, child = "figs_tabs.Rmd"}
```

##### PAGE BREAK
